\MakeUppercase{Linear versus centred Colouring Numbers}\thanks{This research was partly funded by NSERC.}

### 2022-04-05 lin-vs-cen.tex
Pfffft

In this paper, all graphs are simple and undirected. For a graph $G$, $V(G)$ denotes the vertex set of $G$, $E(G)$ denotes the edge set of $G$ and $\VE(G):=V(G)\cup E(G)$.

We make use of the following (polygamous) generalisation of Hall's Marriage Theorem, which follows from Hall's Marriage Theorem by replacing each vertex $x\in X$ with a $d$-tuple $x_1,\ldots,x_d$ each having the same neighbours as $x$.

\begin{cor}\label{d_hall}
  Let $d\ge 1$ be an integer and let $G$ be a bipartite graph with parts $X$ and $Y$ such that, for any $A\subseteq X$, $|N_G(A)|\ge d|A|$.  Then $G$ contains a subgraph $M$ such that $\deg_M(x)=d$ for each $x\in X$ and $\deg_M(y)\le 1$ for each $y\in Y$.
\end{cor}

We will also make use of the following weighted version of the Lovász Local Lemma:

\begin{lem}\label{weighted_lovasz}
  Let $E:=\{E_1,\ldots,E_n\}$ be a set of events in some probability space $(\Omega,\Pr)$.  For each $i\in\{1,\ldots,n\}$, let $\Gamma_i\subseteq E$ be such that the event $E_i$ is mutually independent of $E\setminus \Gamma_i$,\footnote{An event $A$ is mutually independent of a set $\{B_1,\ldots,B_r\}$ of events if, for any disjoint sets $I,J\subseteq\{1,\ldots,r\}$, $\Pr(A\cap\bigcap_{i\in I} B_i\cap\bigcap_{j\in J} \overline{B}_j=\Pr(A)\Pr(\bigcap_{i\in I} B_i\cap\bigcap_{j\in J} \overline{B}_j)$.} and let $w:E\to(0,1)$ be such that
  \[
      \Pr(E_i) \le w(E_i)\cdot\prod_{E_j\in\Gamma_i}(1-w(E_j))  \enspace .
  \]
  Then $\Pr(\overline{E_1}\cap\cdots\cap\overline{E_n}) > 0$.
\end{lem}

% \begin{proof}
%   \todo[inline]{Surely this is known}
% \end{proof}
%
%
% \todo[inline]{Surely this next lemma is well-known?}
% We make use of the following generalization of Hall's Marriage Theorem \cite{hall:on}:
% \begin{lem}\label{hall_vees}
%   Let $G$ be a bipartite graph with vertex parts $A$ and $B$ and having the property that, for any $A_0\subseteq A$, $|N_G(W)|\ge 2|A_0|$.  Then $G$ contains a subgraph $\Lambda$ in which each vertex of $A$ has degree exactly two and each vertex of $B$ has degree at most one.
% \end{lem}
%
% \begin{proof}
%   Let $\Lambda$ be a subgraph of $G$ in which each vertex of $A$ has degree at most two, each vertex of $B$ has degree at most one, and that maximizes $\deg_\Lambda(A)$.  If $\deg_{\Lambda}(A) = 2|\Lambda|$ then there is nothing to prove. Assume therefore, for the sake of contradiction,  that there exists $v_0\in A$ with $\deg_\Lambda(v_0) < 2$.
%
%   We say that a path $v_0,\ldots,v_m$ in $G$ is \emph{$\Lambda$-aternating} if $v_iv_{i+1}\in E(\Lambda)$ for all odd $i\in\{0,\ldots,m-1\}$ and $v_iv_{i+1}\not\in E(\Lambda)$ for all even $i\in\{0,\ldots,m\}$.  Observe that any such path has even length so that it ends at a vertex $v_m\in A$.  Otherwise, replacing the edges $\lfloor m/2\rfloor$ edges $\{v_1v_2,v_3v_4,\ldots,v_{m-2}v_{m-1}\}\in E(\Lambda)$ with the $\lceil m/2\rceil$ edges $\{v_0v_1,v_2v_3,\ldots,v_{m-1}v_m\}$ does not change $\deg_\Lambda(v)$ for any $v\in A_0\setminus\{v_0\}$ but does increase  $\deg_\Lambda(v_0)$, contradicting the assumption that $\Lambda$ maximizes $\deg_\Lambda(A)$.
%
%   Let $A_0\subseteq A$ and $B_0\subseteq B$ be the subsets of $A$ and $B$, respectively, that can be reached from $v_0$ by $\Lambda$-alternating paths.
%   Observe that, for any $x\in B_0$,  there is exactly one edge $vx$ of $\Lambda$ incident on $x$ and the other endpoint $v$ of this edge is in $A_0$.  Let $a$ be the number of edges of $\Lambda$ having one endpoint $A_0$ and one endpoint in $B_0$, so that $a=\deg_\Lambda(B_0)=|B_0|$.  Let $b$ be the number of edges of $\Lambda$ having one endpoint in $A_0$ and one endpoint in $B\setminus B_0$, so that $\deg_\Lambda(A_0)=a+b = |B_0|+b$.    Since $\deg_{\Lambda}(v) \le 2$ for each $v\in A_0$ and $\deg_\Lambda(v_0)<2$, $\deg_\Lambda(A_0)<2|A_0|$.  This yields the desired contradiction, since $|N_\Lambda(A_0)| \le \deg_\Lambda(A_0) = |B_0|+b = \deg_\Lambda(A_0)< 2|A_0|$.greater than
% \end{proof}
%
% \begin{rem}
%   The constant $2$ that appears twice in \cref{hall_vees} can be replaced with any positive integer $c$.
% \end{rem}

@@

% Refer to \cref{grid_figure}.
For positive integers $a$ and $b$, the \emph{$a\times b$ grid} $G_{a\times b}$ is the graph with vertex set $V(G_{a\times b}):=\{1,\ldots,a\}\times\{1,\ldots,b\}$ and that contain an edge with endpoints $(v,w)$ and $(x,y)$ if and only if $|v-x|+|w-y|=1$.  For each $i\in\{1,\ldots,a\}$, the \emph{$i$th column} of $G_{a\times b}$ is the vertex set $\{(i,1),\ldots,(i,b)\}$ and, for each $j\in\{1,\ldots,b\}$, the $j$th row is the vertex set $\{(1,j),\ldots,(a,j)\}$.

In this work, we need to work with pseudogrids.  An \defin{$a\times b$-pseudogrid} is any graph that can be obtained from $G_{a\times b}$ in the following way:
\begin{compactitem}
  \item Replace each edge $vw$ of $G_{a\times b}$ with a path $\overline{P}_{vw}$ whose endpoints are $v$ and $w$.  (In other words, $\overline{P}_{vw}$ is a path obtained by subdividing the edge $vw$ zero or more times.)
  \item Replace each degree-$4$ vertex $(x,y)$ of $G_{a\times b}$ with a path $P_v$. If $P_v$ has only one vertex then the unique vertex of $P_v$ is adjacent to each of $(x\pm 1,y\pm 1)$ and this has effect.  Otherwise, $P_v$ has two endpoints $p$ and $q$, each of which is adjacent to two vertices among $(x\pm 1,y\pm 1)$.  It is useful to consider three possible cases:
  \begin{compactenum}[(Q1)]
    \item \label{q_i} $p$ is adjacent to $\{(x-1,y), (x,y-1)\}$ and $q$ is adjacent to $(x+1,y),(x,y+1)$;
    \item \label{q_ii} $p$ is adjacent to $\{(x,y+1), (x-1,y)\}$ and $q$ is adjacent to $(x,y-1),(x+1,y)$; or
    \item \label{q_iii} $p$ is adjacent to $\{(x-1,y),(x+1,y)\}$ and $q$ is adjacent to $\{(x,y-1),(x,y+1)\}$.
  \end{compactenum}
\end{compactitem}

Let $G$ be an $a\times b$ pseudogrid.  For an edge $vw$ of $G_{a\times b}$, we let $P_{vw}:=\overline{P}_{vw}-\{v,w\}$ denote the (possibly empty) subpath containing the internal vertices of $\overline{P}_{vw}$.  For each vertex $v$ of $G_{a\times b}$ of degree less than $4$ we define $P_{v}$ to be the $1$-vertex path that contains only $v$.  In this way, $\mathcal{P}:=\{V(P_\mu):\mu\in \VE(G_{a\times b})\}$ is a partition of $V(G)$ into induced paths and the quotient graph $G/\mathcal{P}$ is isomorphic to $G_{a\times b}$.  We call $\mathcal{P}$ a \defin{grid-partition} of $G$.
% and we denote it by $(P_\mu:\mu\in \VE(G_{a\times b}))$.
% This defines a mapping $\delta:V(G)\to V(G_{a\times b})\cup E(G_{a\times b})$ where $u\in V(P_{\delta(u)})$ for each $u\in v(G)$.

% Then each vertex $u$ of $G$ naturally corresponds to an edge or vertex $u'$ of $G_{a\times b}$:  If $u$ is a vertex of $P_{vw}$ then $u'=vw$; if $u$ a vertex of $P_v$ then $u'=v$; and otherwise $u'=u$.  This latter case occurs precisely when $u'$ is one of the boundary vertices of $G_{a\times b}$.

@@

We now proceed with the various steps in the proof.

\subsection{Pseudogrids with Exclusively Frequent Colours}

@@

% We extend $\varphi$ to a partial colouring $\phi:V(G_{k\times k})\cup E(G_{k\times k})\to\{1,\ldots,c,\perp\}$ of the edges and vertices of $G_{k\times k}$ as follows:  For each vertex $v$ of $G_{k\times k}$, define $\phi(v)$ be an arbitrary colour in $\{\varphi(x):x\in V(P_v)\}$.  For each edge $vw$ of $G_{k\times k}$, there are two possibilities:
% If $P_{vw}$ is empty then $\phi(vw):=\perp$, otherwise $\phi(vw)$ is an arbitrary colour in $\{\varphi(x):x\in V(P_{vw})\}$.

@@

\begin{proof}
  The proof is by induction on $|\varphi(V(G))|$, the number of colours used by the colouring $\varphi$.  If there exists no $A\subseteq \varphi(V(G))$ with $\varphi_{\mathcal{P}}^{-1}(A) < d|A|$ then taking $G':=G$ and $\mathcal{P}':=\mathcal{P}$ satisfies the requirements of the lemma. Otherwise, we will remove a set $R$ of rows and a set $C$ of columns from $G$ with $|R|=|C|\le d|A|$ to eliminate all vertices with colours in $A$, as follows:
  \begin{compactitem}
    \item For each $v:=(i,j)\in V(G_{k})$ with $A\cap\varphi_\mathcal{P}(v)\neq\emptyset$, we include row $j$ in $R$ and column $i$ in $C$.
    \item For each horizontal edge $vw\in E(G_{k\times k})$ with $A\cap\varphi_\mathcal{P}(vw)\neq\emptyset$, we include the row of $G$ that contains $P_{vw}$ in $R$.
    \item For each vertical edge $vw\in E(G_{k\times k})$ with $A\cap\varphi_\mathcal{P}(vw)\neq\emptyset$, we include the row of $G$ that contains $P_{vw}$ in $R$.\todo{Define horizontal and vertical edges of $G$.}
    \item Finally, we add arbitrary rows to $R$ or columns to $C$ to ensure that $|R|=|C|$.
  \end{compactitem}
  At this point $|R|=|C|\le d|A|$ and we remove all rows in $R$ and all columns in $C$ from $G$ to obtain a $k_0\times k_0$ pseudogrid $G_0$ with $k_0\ge k-d|A|$ and such that $\varphi(V(G_0))\cap A=\emptyset$.  In particular, $|\varphi(V(G_0))|\le |\varphi(V(G))|-|A|$.

@@

\subsection{Finding Paths Through Well-Separated Pairs}

Next we show that, given a sufficiently ``well-separated'' set $S$ of pairs of vertices in $G$, we can always find a path in $G$ that contains every vertex in $S$. For this, we need some definitions of boxes in $G$ and in $G_k$.

@@

Before finishing with this section we recall a fairly standard looking packing lemma.

@@

\todo[inline]{Fix notation conflict:  $G_r(\mu)$ versus $G_k$?  Maybe go back to $G_{k\times k}$?}

\subsection{Finding an Uncentred Path}

@@

  We now want to use the Local Lemma (\cref{weighted_lovasz}) to show that, with probability greater than zero, $\deg_H(\alpha)\in \Omega(d/r^2)$ for each $\alpha\in X$.
  For each $\alpha\in X$, let $\Phi_\alpha:=\{\nu\in\VE(G_k):\phi(\nu)=\alpha\}$ and recall that, since $\phi$ comes from \cref{one_colour_per_object}, $|\Phi_\alpha| \ge d$.  Furthermore all but at most one element of $\Phi_\alpha$ is contained in $\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu))$ since, otherwise the first round would have succeeded for colour $\alpha$.  For each $\mu\in Y$, $|\VE(G_{r+1}(\mu))|\in O(r^2)$. Therefore, there are $\Omega(|\Phi_\alpha|/r^2)$ objects $\mu\in Y$ such that $\VE(G_{r+1}(\mu))\cap \Phi_\alpha\neq\emptyset$.

  We begin by showing that, for each $\alpha\in X$, the random variable $\deg_H(\alpha)$ dominates a $\binomial(\Omega(|\Phi_\alpha|/r^2),1/25)$ random variable, which allows us to establish en exponential inequality for $\Pr(\deg_H(\alpha) < 100r^2)$.  (We make no attempt to optimize the constants here.)  Suppose that, for some $\mu\in Y$, $B_r(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$. Then, by \cref{packing_lemma}, the set
  \[
     Y_\nu := \{\tau\in Y\setminus\{\mu\}:\nu\in\VE(G_r(\tau))\}
  \]
  has size at most $25$. Observe that the edge $\alpha\mu$ will appear in $H$ if, in our random permutation $\pi$ of $Y$, $\nu$ appears before any element of $Y_\nu$.  This happens with probability exactly $1/|Y_\nu|\ge 1/25$.  This already establishes that $\E(\deg_H(\alpha)) \in \Omega(|\Phi_\alpha|/r^2)$.

  In order to obtain a sufficiently strong concentration result for $\deg_H(\alpha)$ we need to find some independence.  To do this, consider the graph $J$ with vertex set $V(J)=Y$ and which contains an edge $\mu_1\mu_2$ if $B_{2r}(\mu_1)\cap B_{2r}(\mu_2)\neq\emptyset$.  Now, consider the greedy process of repeatedly choosing a vertex $\mu$ of $J$ such that $B_r(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$ and then set $J:=J-\VE(G_{10r}(\mu))$.  This process continues until no suitable vertex of $J$ remains.  Each step in this process eliminates $O(r^2)$ objects $\nu$ with $\varphi(\nu)=\alpha$,\todo{Explain better} so the number of iterations is $r\in\Omega(d/r^2)$.

  Let $\{\mu_1',\ldots,\mu_t'\}$ be the subset of $Y$ chosen by this process and let $\{\nu_1,\ldots,\nu_r\}\subseteq\Phi_\alpha$ be objects with $\nu_i\in\VE(G_r(\mu_i'))$, for each $i\in\{1,\ldots,t\}$.  The important observation now is that the sets $X_{\nu_1},\ldots,X_{\nu_t}$ are disjoint.  Therefore, if we let $U_i$ denote the event ``$\mu_i'$ apppears before any elements of $X_{\nu_i}$ in our random permutation of $Y$'' then the events $U_1,\ldots,U_k$ are mutually independent.  Indeed, each $U_i$ depends only on the relative order of $\{\mu_i'\}\cup X_{\nu_i}$ in the random permutation of $Y$.  For each $i\in\{1,\ldots,k\}$, $\Pr(U_i)\ge 1/25$.  Therefore, $\sum_{i=1}^k \mathbbm{1}_{U_i}$ dominates a $\binomial(k,1/25)$ random variable.  Therefore,
  \begin{align*}
    \Pr\left(\deg_H(\alpha)< 100r^2\right)
    & \le \Pr\left(\sum_{i=1}^t \mathbbm{1}_{U_i} < 100r^2\right) \\
    & \le \Pr(\binomial(t,1/25)\le 100r^2) \\
    & \le (2t)^{100r^2}(1/25)^{r^2}(1-1/25)^{t-100r^2} \\
    & \le \exp(100r^2\log(2t)+(t-100r^2)\log(1-1/?)) \\
    & \le \exp(-100r^2\log(2t)\log(1-1/?) + t\log(1-1/25)) \\
    & = \exp(ar^2 - bt)
  \end{align*}
  for fixed positive $a,b>0$.  Recall that we are working in the regime in  $t\in\Omega(|\Phi_\alpha|/r^2)$ and $|\Phi_\alpha|\in\Omega(d/r^2)$ so, for any fixed $C>0$, there exists $d\in\Omega(r^4)$ such that
  $\Pr(\deg_H(\alpha)< 100r^2)\le \exp(-C|\Phi_\alpha|/r^2)$.
  \todo[inline]{Work the preceding calculation out}
  We are now ready for an application of \cref{weighted_lovasz}.\todo{Continue here}  For each $\alpha\in X$, let $E_\alpha$ denote the event ``$\deg_H(\alpha)\le 100r^2$.''  For each $\alpha\in X$, define $\Gamma_\alpha$ to be the set that contains every event $E_\beta$ such that $\Phi_\alpha$ contains an element $\nu$ with $B_{10r}(\nu)\cap Z_\beta\neq\emptyset$.  The event $X_\alpha$ is mutually independent of $\{X_\beta:\beta\in X\setminus\{\alpha\}\setminus\Gamma_\alpha\}$ since, for any $E_\beta\not\in \Gamma_\alpha$ the sets $Y_\alpha:=\{\mu\in Y:B_{2r}(\mu)\cap \Phi_\alpha\neq\emptyset\}$ and $Y_\beta:=\{\mu\in Y:B_{2r}(\mu)\cap Z_\beta\neq\emptyset\}$ are disjoint and the event $E_\alpha$ is entirely determined by the relative order of $Y_\alpha$ in the permutation while $E_\beta$ is entirely determined by the relative order of $Y_\beta$.\todo{Explain better}   Finally, observe that $|\Gamma_\alpha|\le 25|\Phi_\alpha|$.

@@

>>>>

In this paper, all graphs are simple and undirected. For a graph $G$, $V(G)$ denotes the vertex set of $G$, $E(G)$ denotes the edge set of $G$ and $\VE(G):=V(G)\cup E(G)$ denotes the set of vertices and edges of $G$. For a vertex $v\in V(G)$ we use the notations $N_G(v):=\{w\in V(G):vw\in E(G)\}$ and $N_G[H]:=N_G(H)\cup\{v\}$ to denote the open and close neighbourhoods of $v$ in $G$.  For a set $S\subseteq V(G)$, $N_G(S):=\{w\in V(G):vw\in E(G),\, v\in S,\, w\not\in S\}$ and $N_G[S]:=\bigcup_{v\in S} N_G[v]$.

For a  bipartite graph $H$, the two parts of $V(H)$ are denoted by $L(H)$ and $R(H)$ and we use the convention that an edge $vw$ has $v\in L(H)$ and $w\in R(H)$.  A \defin{matching} $M$ in a bipartite graph $H$ is a subgraph of $H$ in which each vertex has degree at most $M$.  We say that $M$ \defin{saturates} a set $S\subseteq V(H)$ if $\deg_M(v)=1$ for each $v\in S$.  We make use of Hall's Famous Marriage Theorem (see, for example \citet[Theorem~2.1.2]{diestel:graph}):

\begin{lem}\label{hall}
  Let $H$ be a bipartite graph with the property that $|N_H(A)|\ge |A|$ for each $A\subseteq L(H)$.  Then $H$ contains a matching that saturates $L(H)$.
\end{lem}

More specifically, we make use of the following (polygamous) generalisation of \cref{hall}, which follows from \cref{hall} by replacing each vertex $x\in L(H)$ with $d$ vertices $x_1,\ldots,x_d$ each having the same neighbourhood as $x$.  More precisely, we apply \cref{hall} on the graph $H':=H-\{x\}\cup\{x_iy:i\in\{1,\ldots,d\},\, y\in N_H(x)\}$.

\begin{cor}\label{d_hall}
  Let $d\ge 1$ be an integer and let $H$ be a bipartite graph with the property that $|N_H(A)|\ge d|A|$ for each $A\subseteq L(H)$.  Then $H$ contains a subgraph $M$ such that $\deg_M(v)=d$ for each $v\in L(H)$ and $\deg_M(w)\le 1$ for each $y\in R(H)$.
\end{cor}

Late in the game, we will make use of the following asymmetric version of the Lovász Local Lemma (see, for example, \citet[Lemma~5.1.1]{alon.spencer:probabilistic}):

\begin{lem}\label{weighted_lovasz}
  Let $\mathcal{E}:=\{E_1,\ldots,E_n\}$ be a set of events in some probability space $(\Omega,\Pr)$.  For each $i\in\{1,\ldots,n\}$, let $\Gamma_i\subseteq \mathcal{E}$ be such that the event $E_i$ is mutually independent of $\mathcal{E}\setminus \Gamma_i$,\footnote{An event $A$ is mutually independent of a set $\{B_1,\ldots,B_r\}$ of events if, for any disjoint sets $I,J\subseteq\{1,\ldots,r\}$, $\Pr(A\cap\bigcap_{i\in I} B_i\cap\bigcap_{j\in J} \overline{B}_j=\Pr(A)\Pr(\bigcap_{i\in I} B_i\cap\bigcap_{j\in J} \overline{B}_j)$.} and let $w:E\to[0,1)$ be such that
  \[
      \Pr(E_i) \le w(E_i)\cdot\prod_{E_j\in\Gamma_i}(1-w(E_j))  \enspace .
  \]
  Then $\Pr(\overline{E}_1\cap\cdots\cap\overline{E}_n) > 0$.
\end{lem}

@@

For positive integers $a$ and $b$, the \emph{$a\times b$ grid} $G_{a\times b}$ is the graph with vertex set $V(G_{a\times b}):=\{1,\ldots,a\}\times\{1,\ldots,b\}$ and that contain an edge with endpoints $(v,w)$ and $(x,y)$ if and only if $|v-x|+|w-y|=1$.  Such an edge is \defin{vertical} if $v=x$ and \defin{horizontal} if $w=y$.  For each $i\in\{1,\ldots,a\}$, the \emph{$i$th column} of $G_{a\times b}$ is the vertex set $\{(i,1),\ldots,(i,b)\}$ and, for each $j\in\{1,\ldots,b\}$, the $j$th row is the vertex set $\{(1,j),\ldots,(a,j)\}$.

\begin{figure}
  \begin{center}
    \includegraphics{figs/pseudogrid-2}
  \end{center}
  \caption{A $6\times 4$ pseudogrid.  Black vertices are contained in $P_v$ for some vertex $v$ of $G_{6\times 4}$ and white vertices are contained in $P_{vw}$ for some edge of $G_{6\times 4}$.}
  \label{pseudogrid_fig}
\end{figure}

In this work, we need to work with pseudogrids.  Refer to \cref{pseudogrid_fig}.  An \defin{$a\times b$-pseudogrid} is any graph that can be obtained from $G_{a\times b}$ in the following way:
\begin{compactitem}
  \item Replace each edge $vw$ of $G_{a\times b}$ with a path $\overline{P}_{vw}$ whose endpoints are $v$ and $w$.  (In other words, $\overline{P}_{vw}$ is a path obtained by subdividing the edge $vw$ zero or more times.)
  \item Replace each degree-$4$ vertex $(x,y)$ of $G_{a\times b}$ with a path $P_v$. If $P_v$ has only one vertex then the unique vertex of $P_v$ is adjacent to each of $(x\pm 1,y\pm 1)$ and this has effect.  Otherwise, $P_v$ has two endpoints $p$ and $q$, each of which is adjacent to two vertices among $(x\pm 1,y\pm 1)$.  It is useful to consider three possible cases:
  \begin{compactenum}[(Q1)]
    \item \label{q_i} $p$ is adjacent to $\{(x-1,y), (x,y-1)\}$ and $q$ is adjacent to $(x+1,y),(x,y+1)$;
    \item \label{q_ii} $p$ is adjacent to $\{(x,y+1), (x-1,y)\}$ and $q$ is adjacent to $(x,y-1),(x+1,y)$; or
    \item \label{q_iii} $p$ is adjacent to $\{(x-1,y),(x+1,y)\}$ and $q$ is adjacent to $\{(x,y-1),(x,y+1)\}$.
  \end{compactenum}
\end{compactitem}

Let $G$ be an $a\times b$ pseudogrid.  For an edge $vw$ of $G_{a\times b}$, we let $P_{vw}:=\overline{P}_{vw}-\{v,w\}$ denote the (possibly empty) subpath containing the internal vertices of $\overline{P}_{vw}$.  For each vertex $v$ of $G_{a\times b}$ of degree less than $4$ we define $P_{v}$ to be the $1$-vertex path that contains only $v$.  In this way, $\mathcal{P}:=\{V(P_\mu):\mu\in \VE(G_{a\times b})\}$ is a partition of $V(G)$ into induced paths and the quotient graph $G/\mathcal{P}$ is isomorphic to $G_{a\times b}$.  We call $\mathcal{P}$ a \defin{grid-partition} of $G$.

@@

In \cref{only_frequent_sec} we explain how to remove rows and columns of $G$ to obtain a sub-pseudogrid in which every colour appears frequently.  In \cref{path_finding} we explain how, given a well-separated set $S$ of vertices in $G$, to find a path that contains every vertex in $S$.  In \cref{packing_lemma_sec} we introduce fairly standard-looking packing lemma that is needed in several places later.  In \cref{well_separated} we explain how find a well-separated set $S$ that contains two vertices of each colour. Finally \cref{wrapping_up} shows how to assemble these various pieces to prove \cref{pseudogrid_lower_bound}.

\subsection{Pseudogrids with Exclusively Frequent Colours}
\label{only_frequent_sec}

@@

@@

\begin{proof}
  The proof is by induction on $|\varphi(V(G))|$, the number of colours used by the colouring $\varphi$.  If there exists no $A\subseteq \varphi(V(G))$ with $\varphi_{\mathcal{P}}^{-1}(A) < d|A|$ then taking $G':=G$ and $\mathcal{P}':=\mathcal{P}$ satisfies the requirements of the lemma. Otherwise, we will remove a set $R$ of rows and a set $C$ of columns from $G$ with $|R|=|C|\le d|A|$ to eliminate all vertices with colours in $A$, as follows:
  \begin{compactitem}
    \item For each $v:=(i,j)\in V(G_{k})$ with $A\cap\varphi_\mathcal{P}(v)\neq\emptyset$, we include row $j$ in $R$ and column $i$ in $C$.
    \item For each horizontal edge $vw\in E(G_{k\times k})$ with $A\cap\varphi_\mathcal{P}(vw)\neq\emptyset$, we include the row of $G$ that contains $P_{vw}$ in $R$.
    \item For each vertical edge $vw\in E(G_{k\times k})$ with $A\cap\varphi_\mathcal{P}(vw)\neq\emptyset$, we include the row of $G$ that contains $P_{vw}$ in $R$.
    \item Finally, we add arbitrary rows to $R$ or columns to $C$ to ensure that $|R|=|C|$.
  \end{compactitem}
  At this point $|R|=|C|\le d|A|$ and we remove all rows in $R$ and all columns in $C$ from $G$ to obtain a $k_0\times k_0$ pseudogrid $G_0$ with $k_0\ge k-d|A|$ and such that $\varphi(V(G_0))\cap A=\emptyset$.  In particular, $|\varphi(V(G_0))|\le |\varphi(V(G))|-|A|$.

@@

\subsection{Finding Paths Through Well-Separated Pairs}
\label{path_finding}

Next we show that, given a sufficiently `well-separated' set $S$ of pairs of vertices in $G$, we can always find a path in $G$ that contains every vertex in $S$. For this, we need some definitions of boxes in $G$ and in $G_k$.

@@

\subsection{A Packing Lemma}
\label{packing_lemma_sec}

We will make use of the following fairly standard looking packing lemma.

@@

\subsection{Finding a Well-Separated Set}
\label{well_separated}

@@

  We now want to use the Local Lemma (\cref{weighted_lovasz}) to show that, with probability greater than zero, $\deg_H(\alpha)\in \Omega(d/r^2)$ for each $\alpha\in X$.
  For each $\alpha\in X$, let $\Phi_\alpha:=\{\nu\in\VE(G_k):\phi(\nu)=\alpha\}$ and recall that, since $\phi$ comes from \cref{one_colour_per_object}, $|\Phi_\alpha| \ge d$.  Furthermore all but at most one element of $\Phi_\alpha$ is contained in $\bigcup_{\mu\in Y}\VE(G_{r+1}(\mu))$ since, otherwise the first round would have succeeded for colour $\alpha$.
  % For each $\mu\in Y$, $|\VE(G_{r+1}(\mu))|\in O(r^2)$. Therefore, there are $\Omega(|\Phi_\alpha|/r^2)$ objects $\mu\in Y$ such that $\VE(G_{r+1}(\mu))\cap \Phi_\alpha\neq\emptyset$.

  We begin by showing that, for each $\alpha\in X$, the random variable $\deg_H(\alpha)$ dominates a $\binomial(\Omega(|\Phi_\alpha|/r^2),1/25)$ random variable, which allows us to establish en exponential inequality for $\Pr(\deg_H(\alpha) < 100r^2)$.  (We make no attempt to optimize the constants here.)  Suppose that, for some $\mu\in Y$, $B_r(\mu)\cap \Phi_\alpha$ contains at least one element $\nu$. Then, by \cref{packing_lemma}, the set
  \[
     Y_\nu := \{\tau\in Y:\nu\in\VE(G_{2r}(\tau))\}
  \]
  has size at most $25$. Observe that the edge $\alpha\mu$ will appear in $H$ if, in our random permutation $\pi$ of $Y$, $\nu$ appears before any other element of $Y_\nu$.  This happens with probability exactly $1/|Y_\nu|\ge 1/25$.
   % This already establishes that $\E(\deg_H(\alpha)) \in \Omega(|\Phi_\alpha|/r^2)$.

  In order to obtain a sufficiently strong concentration result for $\deg_H(\alpha)$ we need to find some independence.  To do this, set $J:=\Phi_\alpha$ and consider the greedy process of repeatedly choosing some $\mu\in Y$ such that $\VE(G_r(\mu))\cap J$ contains at least one element $\nu$ and then set $J:=J-\VE(G_{10r}(\mu))$.\footnote{The constant $10$ is overkill here and is only used for simplicity; $8r+O(1)$ would be sufficient.}  This process continues until $B_r(\mu)\cap J=\emptyset$ for each $\mu\in Y$. Since $|\VE(G_{10r}(\mu))|\in O(r^2)$, each step in this process eliminates $O(r^2)$ objects from $J$ so the number $t$ of iterations is in $\Omega(|\Phi_\alpha|/r^2)\subseteq \Omega(d/r^2)$.

  Let $\{\mu_1',\ldots,\mu_t'\}$ be the subset of $Y$ chosen by this process and let $\{\nu_1,\ldots,\nu_r\}\subseteq\Phi_\alpha$ be the corresponding elements of $\Phi_\alpha$.  The important observation now is that the sets $Y_{\nu_1},\ldots,Y_{\nu_t}$ are disjoint.  Therefore, if we let $U_i$ denote the event ``$\mu_i'$ apppears in $\pi$ before any other element of $Y_{\nu_i}$'' then the events $U_1,\ldots,U_k$ are mutually independent.  Indeed, each $U_i$ depends only on the relative order of $Y_{\nu_i}$ in the random permutation of $Y$.  For each $i\in\{1,\ldots,k\}$, $\Pr(U_i)\ge 1/25$.  Therefore, $\sum_{i=1}^t \mathbbm{1}_{U_i}$ dominates a $\binomial(t,1/25)$ random variable.  Therefore,
  \begin{align*}
    \Pr\left(\deg_H(\alpha)< 100r^2\right)
    & \le \Pr\left(\sum_{i=1}^t \mathbbm{1}_{U_i} < 100r^2\right) \\
    & \le \Pr(\binomial(t,1/25)\le 100r^2) \\
    & \le (2t)^{100r^2}(1/25)^{r^2}(1-1/25)^{t-100r^2} \\
    & \le \exp(100r^2\log(2t)+(t-100r^2)\log(1-1/?)) \\
    & \le \exp(-100r^2\log(2t)\log(1-1/?) + t\log(1-1/25)) \\
    & = \exp(ar^2 - bt)
  \end{align*}
  for fixed positive $a,b>0$.  Recall that we are working in the regime in  $t\in\Omega(|\Phi_\alpha|/r^2)$ and $|\Phi_\alpha|\in\Omega(d/r^2)$ so, for any fixed $C>0$, there exists $d\in\Omega(r^4)$ such that
  $\Pr(\deg_H(\alpha)< 100r^2)\le \exp(-C|\Phi_\alpha|/r^2)$.\todo{This could be done better.} Let $E_\alpha$ denote the event ``$\deg_H(\alpha)\le 100r^2$.''  By the preceding discussion, $\Pr(E_\alpha)\le \exp(-C|\Phi_\alpha|/r^2)$.

  We are now ready for an application of \cref{weighted_lovasz}.  Let $\mathcal{E}:=\{E_\alpha:\alpha\in X\}$ and, for each $\alpha\in X$, define
  \[
     \Gamma_\alpha : \left\{E_\beta \in \mathcal{E}: \Phi_\beta\cap \bigcup_{\nu\in \Phi_\alpha} \VE(G_{10r}(\nu))\neq\emptyset\right\} \enspace .
  \]
  The event $E_\alpha$ is mutually independent of $\mathcal{E}\setminus\Gamma_\alpha$ since, for any $E_\beta\in\mathcal{E}\setminus\Gamma_\alpha$ the sets $Y_\alpha:=\{\mu\in Y:B_{2r}(\mu)\cap \Phi_\alpha\neq\emptyset\}$ and $Y_\beta:=\{\mu\in Y:B_{2r}(\mu)\cap \Phi_\beta\neq\emptyset\}$ are disjoint and $\deg_H(\alpha)$ and $\deg_H(\beta)$ are each determined by the relative orders of $Y_\alpha$ and $Y_\beta$ in $\pi$, respectively.

  For each $\alpha\in X$, let $w(E_\alpha):=1-e^{-C/50r^2}$.  Then
  \begin{align}
    w(E_\alpha)\prod_{E_\beta\in\Gamma_\alpha} (1-w(E_\beta))
    & = (1-e^{-C/50r^2})e^{-C|\Gamma_\alpha|/50r^2} \\
    & = e^{-C|\Gamma_\alpha|/50r^2}-e^{-C|(\Gamma_\alpha|+1)/50r^2} \\
    & \ge e^{-C|\Phi_\alpha|/2r^2}-e^{-C|(\Gamma_\alpha|+1)/50r^2} \enspace . \label{ratioo}
    % & \ge (1-e^{-C/50r^2})e^{-C|\Phi_\alpha|/2r^2} \\
    % & \ge (1-e^{-C/50r^2})e^{-C|\Phi_\alpha|/2r^2} \\
  \end{align}
  Taking the ratio of \cref{ratioo} and $\Pr(E_\alpha)$ we obtain
  \begin{align*}
    \frac{e^{-C|\Phi_\alpha|/2r^2}-e^{-C|(\Gamma_\alpha|+1)/50r^2}}{\Pr(E_\alpha)}
    & \ge \frac{e^{-C|\Phi_\alpha|/2r^2}-e^{-C|(\Gamma_\alpha|+1)/50r^2}}{e^{-C|\Phi_\alpha|/r^2}} \\
    & \ge e^{C|\Phi_\alpha|/2r^2} - \frac{e^{-C|(\Gamma_\alpha|+1)/50r^2}}{e^{-C|\Phi_\alpha|/r^2}} \\
  \end{align*}

  \todo[inline]{Hmmmmm. Not working.  Try something like $w(E_\alpha)=1-e^{-1/50}$}

  Finally, by \cref{packing_lemma} $|\Gamma_\alpha|\le 25|\Phi_\alpha|$.

@@

\subsection{Wrapping Up}
\label{wrapping_up}
